On the instability and degeneracy of deep learning models

### 2017-11-07 writing/paper-note/chapters/instab.Rmd
Vardeman's edits, submitted to tech

We consider the behavior, and the potential impropriety, of sequences of discrete probability models built to incorporate observations of increasing sample size $N$. Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.  The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by @schweinberger2011instability for a family of exponential models. At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation. As an example, such models can include near-degenerate distributions that assign all essential probability mass to only a subset of an overall sample space. The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes  may not resemble observed data [cf. @handcock2003assessing]. As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions. Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate our general study of instability for a broad class of probability models, described next.

@@

Section \@ref(examples) provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models). These are used as references for later illustrations. Section \@ref(instability-results) then establishes several formal results for  FOES models with regard to instability.  @schweinberger2011instability originally developed instability results specific to a certain class of discrete exponential models. For similar exponential models with random networks, @handcock2003assessing studied model degeneracy, where a probability model places near complete mass on modes and may thereby narrow the feasible model outcomes. As findings here and from @schweinberger2011instability suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability. Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class, that incorporates results of @schweinberger2011instability as a special case. We prescribe a general and simple condition for identifying instability in a FOES model sequence, which  quantifies whether certain maximal probabilities in a FOES model  are too extreme relative to the sample size $N$. When fulfilled, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes. As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy. Section \@ref(implications) then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and   MCMC-based simulations. As one potential remedy, suggestions are given or constraining model parameterizations to avoid the most problematic regions of the parameter space. Proofs of the main results appear in Appendix \@ref(appendix-instab).

@@

As a basic characteristic, S-unstable FOES model sequences  have extremely sensitive probability structures. One aspect is that small changes in data configuration can   lead to very large changes in probability. Consider, for example, the quantity given by
$$
\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
which represents the biggest log-probability ratio for a one component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.  We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for S-unstable FOES models.

```{theorem, label="instab-elpr", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models.
\begin{enumerate}[(i)]
\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}, then
    $$ \DN(\boldsymbol \theta_N) > C,$$
    or probabilities from a one component change in some outcome have log-ratio exceeding $C$.
\item Suppose the FOES model sequence is S-unstable. Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
    $$
    \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C].
    $$
\end{enumerate}
```

Theorem \@ref(thm:instab-elpr)(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$. If so, Theorem \@ref(thm:instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability. Furthermore, and perhaps more seriously under Theorem \@ref(thm:instab-elpr)(ii),
S-unstable models can never have universally bounded changes in probability among single component variations in data configurations. While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property. As a consequence, unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.

@@

```{proposition, label="prop1", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, $N \geq 1$,  denote linear exponential models (\ref{eq:expo}) with   parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k$ and statistics  $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k$, for fixed $k \geq 1$. Then, the models $P_{\boldsymbol \theta_N}$ are {\rm S-stable}   if
\begin{equation}
(\#eq:prop1)
\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
\end{equation}
holds, i.e.,  if $\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N$ is bounded sequence of sample size $N$.
```

@@

of outcomes $\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}$, $\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}})  \in\{\pm1\}^{N_{\mathcal{H}}}$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})$, with $\theta_{i,N}^{\mathcal{V}}$, $\theta_{j,N}^{\mathcal{H}}$ and $\theta_{ij,N}^{\mathcal{VH}}$ denoting respective parameter components, $1 \leq i \leq N$, $1 \leq j \leq N_{\mathcal{H}}$. Due to the marginalization steps in defining the distribution \@ref(eq:RBM2) of $\boldsymbol X$, note that $\REP_{\boldsymbol X}(\boldsymbol \theta_N)$ has no immediate analytical expression similar to that of  $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)$. For clarity, recall also that S-instability \@ref(eq:Sun) in each model type refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty$) upon scaling by the corresponding number of variables in a distribution. In the following, let $|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|$ denote the L-1 norm of a generic vector $\boldsymbol y =(y_1,\ldots,y_d)$, $d \geq 1$.

@@

```{r rbm-plots, fig.show='hold', fig.cap="The sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/N_{\\mathcal{V}}$ (left) and $\\DN(\\boldsymbol \\theta)$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of these metrics, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
elpr_summary %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), N[italic(V)]))), low = "yellow", high = "red") +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", N[italic(H)], "*", N[italic(V)], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom")

# fix labels
mean_max_prob_diff_summary %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = mean_max_prob)) +
  geom_contour(aes(x = r1, y = r2, z = mean_max_prob), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", Delta[N](theta))), low = "yellow", high = "red") +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", N[italic(H)], "*", N[italic(V)], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom") 
```

In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters $\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})$ and interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$) to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})$ and $||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})$ for a RBM with $N_\mathcal{V} = `r V`$ visibles and $N_\mathcal{H} = `r H`$ hiddens. These average magnitudes vary on a grid between `r min(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))` and `r max(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))` with `r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))` breaks, yielding `r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))^2` grid points. At each point in the grid, `r n` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and `r n` vectors ($\boldsymbol \theta_{interaction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables. These vectors are then paired to create `r n` values of $\boldsymbol \theta_N$ with magnitudes at each point in the grid. The values $\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}$ and $\DN(\boldsymbol \theta_N)$ are then calculated for each $\boldsymbol \theta_N$ and then summarized for each point in the grid using the sample mean. The results of this numerical study are shown in Figure \@ref(fig:rbm-plots). From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one component change in data outcomes are both increasing away from $\boldsymbol \theta_N = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).

@@

Due to their extreme sensitivity in probability structure, S-instability in FOES models may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods. We describe these aspects in Sections \@ref(mcmc)-\@ref(bayes) with regard to data simulation, maximum likelihood estimation and Bayes inference, respectively.

@@

Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed with $P_{\boldsymbol \theta_N}$ as the stationary distribution. For an unstable FOES model, one component changes in outcomes may produce
radically different probabilities, which may then impose numerical barriers to MCMC. For example, consider implementation of a Gibbs sampler from the full conditional distributions $P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})$, $x\in\mathcal{X}$, of each variable $X_i$ based on values $\boldsymbol x_{-i}\in \mathcal{X}^{N-1}$ for the remaining variables, say $\boldsymbol X_{-i}$, in $\boldsymbol X$. If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\boldsymbol \theta_N}$, then the Gibbs sampler can have extreme log-ratios in its transition probabilities, 
$$
\log \left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|,
$$  
as conditional probabilities are proportional to joint probabilities that, with unstable models, can have unbounded log-probability ratios in one-component changes (Theorem
\@ref(thm:instab-elpr)). This can hinder the ability of chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may mix poorly by moving rapidly to, and slowly away from, sections of the sample space. In this case, for example, the Markov chain may become entrapped within a mode of the probability function, with rare chance  of escaping to adequately mimic the occupation frequencies in the overall sample space. If modes of the unstable model are not unique, then important outcomes may be missed without multiple chains or impractically enormous numbers of MCMC samples. This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in any particulars of the MCMC algorithm, and similar complications can also arise for Metropolis-Hastings algorithms for MCMC. Hence, while an unstable FOES model $P_{\boldsymbol \theta_N}$ may be valid and technically open to simulation by MCMC, the aspect of instability can render applications of MCMC as numerically infeasible for simulation purposes. This result is in line with conclusions of @handcock2003assessing and @schweinberger2011instability for other exponential models.

@@

The potential numerical difficulties with maximum likelihood with unstable models, as described in the previous section, can naturally carry over to Bayes inference. Considering that the degeneracy issues related to unstable models can cause likelihoods can be flat (e.g., near zero) for many parameters under a given data outcome and that sign changes in parameters can shift tremendous probability to extreme and opposite outcomes in the sample space (e.g., Corollary \@ref(cor:sign), Theorem \@ref(thm:something)), then numerical complications may arise with Bayes inference in sampling a posterior parameter space based on MCMC. The potential challenges in chain mixing are similar to those presented in Section \@ref(mcmc), though in chain movements through the parameter space as opposed to the sample space in data generation. That is, in the Bayes setting for sampling a posterior distribution for $\boldsymbol \theta_N$,  a chain may  unstable to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one component changes in the sample space may impact data simulation with unstable models. For example, if $\pi(\cdot)$ denotes a prior density for $\boldsymbol \theta_N$ and $q(\cdot | \cdot)$  denotes a proposal distribution  for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes 
$$
\alpha\left(\boldsymbol \theta^{(1)} \mid \boldsymbol \theta^{(2)}\right)= \min\left\{1,
\frac{q(\boldsymbol \theta_N^{(2)} \mid \boldsymbol \theta_N^{(1)})}{q(\boldsymbol \theta_N^{(1)} \mid \boldsymbol \theta_N^{(2)})}
\frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
$$
which indicates how parameter sensitivity in the likelihood $P_{\boldsymbol \theta_N} (  \boldsymbol x)$ may complicate sampling of the posterior $P_{\boldsymbol \theta_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N)$ (i.e., moving from $\boldsymbol \theta^{(1)}$ to $\boldsymbol \theta^{(2)}$ in the parameter space). Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \@ref(rbm)-\@ref(deep-learning). As latent variables are often sampled with parameters in a Bayes MCMC approach, this aspect may further compound numerical problems in chain mixing.

>>>>

We consider the behavior, and the potential impropriety, of sequences of discrete probability models built to incorporate observations of increasing sample size $N$. Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.  The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by @schweinberger2011instability for a family of exponential models. At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation. As an example, such models can include near-degenerate distributions that assign essentially all probability mass to only a subset of an overall sample space. The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes may not resemble observed data [cf. @handcock2003assessing]. As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions. Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate our general study of instability for a broad class of probability models, described next.

@@

Section \@ref(examples) provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models). These are used as references for later illustrations. Section \@ref(instability-results) then establishes several formal results for  FOES models with regard to instability.  @schweinberger2011instability originally developed instability results specific to a certain class of discrete exponential models. For similar exponential models with random networks, @handcock2003assessing studied model degeneracy, where a probability model places near complete mass on modes and may thereby narrow the feasible model outcomes. As findings here and from @schweinberger2011instability suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability. Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class, that incorporates results of @schweinberger2011instability as a special case. We prescribe a general and simple condition for identifying instability in a FOES model sequence, which  quantifies whether certain maximal probabilities in a FOES model  are too extreme relative to the sample size $N$. When these conditions are met, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes. As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy. Section \@ref(implications) then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and MCMC-based simulations. As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space. Proofs of the main results appear in Appendix \@ref(appendix-instab).

@@

As a basic characteristic, S-unstable FOES model sequences  have extremely sensitive probability structures. One aspect is that small changes in data configuration can   lead to very large changes in probability. Consider, for example, the quantity given by
$$
\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
which represents the biggest log-probability ratio for a one-component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.  We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for S-unstable FOES models.

```{theorem, label="instab-elpr", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models.
\begin{enumerate}[(i)]
\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}), then
    $$ \DN(\boldsymbol \theta_N) > C,$$
    or probabilities from a one-component change in some outcome have log-ratio exceeding $C$.
\item Suppose the FOES model sequence is S-unstable. Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
    $$
    \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C].
    $$
\end{enumerate}
```

Theorem \@ref(thm:instab-elpr)(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$. If so, Theorem \@ref(thm:instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability. Furthermore, and perhaps more seriously under Theorem \@ref(thm:instab-elpr)(ii),
S-unstable models can never have universally bounded changes in probability among single component variations in data configurations. While not all one-component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property. As a consequence, unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.

@@

```{proposition, label="prop1", echo=TRUE}
Let $P_{\boldsymbol \theta_N}$, $N \geq 1$,  denote linear exponential models (\ref{eq:expo}) with   parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k$ and statistics  $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k$, for fixed $k \geq 1$. Then, the models $P_{\boldsymbol \theta_N}$ are S-stable if
\begin{equation}
(\#eq:prop1)
\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
\end{equation}
holds, i.e.,  if $\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N$ is bounded sequence of sample size $N$.
```

@@

of outcomes $\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}$, $\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}})  \in\{\pm1\}^{N_{\mathcal{H}}}$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})$, with $\theta_{i,N}^{\mathcal{V}}$, $\theta_{j,N}^{\mathcal{H}}$ and $\theta_{ij,N}^{\mathcal{VH}}$ denoting respective parameter components, $1 \leq i \leq N$, $1 \leq j \leq N_{\mathcal{H}}$. Due to the marginalization steps in defining the distribution \@ref(eq:RBM2) of $\boldsymbol X$, note that $\REP_{\boldsymbol X}(\boldsymbol \theta_N)$ has no immediate analytical expression similar to that of  $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)$. For clarity, recall also that S-instability \@ref(eq:Sun) in each model type refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty$) upon scaling by the corresponding number of variables in a distribution. In the following, let $|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|$ denote the L1 norm of a generic vector $\boldsymbol y =(y_1,\ldots,y_d)$, $d \geq 1$.

@@

```{r rbm-plots, fig.show='hold', fig.cap="The sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/N_{\\mathcal{V}}$ (left) and $\\DN(\\boldsymbol \\theta)$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of these metrics, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
elpr_summary %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), N[italic(V)]))), low = "white", high = "grey20") +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", N[italic(H)], "*", N[italic(V)], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom")

# fix labels
mean_max_prob_diff_summary %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = mean_max_prob)) +
  geom_contour(aes(x = r1, y = r2, z = mean_max_prob), colour = "black", bins = 8) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(paste("Mean ", Delta[N](theta))), low = "white", high = "grey20") +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", N[italic(H)], "*", N[italic(V)], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom")
```

In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters $\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})$ and interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$) to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})$ and $||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})$ for a RBM with $N_\mathcal{V} = `r V`$ visibles and $N_\mathcal{H} = `r H`$ hiddens. These average magnitudes vary on a grid between $`r min(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))`$ and $`r max(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))`$ with $`r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))`$ breaks, yielding $`r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))^2`$ grid points. At each point in the grid, $`r n`$ vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and $`r n`$ vectors ($\boldsymbol \theta_{interaction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables. These vectors are then paired to create $`r n`$ values of $\boldsymbol \theta_N$ with magnitudes at each point in the grid. The values $\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}$ and $\DN(\boldsymbol \theta_N)$ are then calculated for each $\boldsymbol \theta_N$ and then summarized for each point in the grid using the sample mean. The results of this numerical study are shown in Figure \@ref(fig:rbm-plots). From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one-component change in data outcomes are both increasing away from $\boldsymbol \theta_N = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).

@@

Due to their induced sensitivity in probability structure, S-instability in FOES models may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods. We describe these aspects in Sections \@ref(mcmc)-\@ref(bayes) with regard to data simulation, maximum likelihood estimation and Bayes inference, respectively.

@@

Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed with $P_{\boldsymbol \theta_N}$ as the stationary distribution. For an unstable FOES model, one-component changes in outcomes may produce
radically different probabilities, which may then impose numerical barriers to MCMC. For example, consider implementation of a Gibbs sampler from the full conditional distributions $P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})$, $x\in\mathcal{X}$, of each variable $X_i$ based on values $\boldsymbol x_{-i}\in \mathcal{X}^{N-1}$ for the remaining variables, say $\boldsymbol X_{-i}$, in $\boldsymbol X$. If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\boldsymbol \theta_N}$, then the Gibbs sampler can have extreme log-ratios in its transition probabilities, 
$$
\log \left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|,
$$  
as conditional probabilities are proportional to joint probabilities that, with unstable models, can have unbounded log-probability ratios in one-component changes (Theorem
\@ref(thm:instab-elpr)). This can hinder the ability of a chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may mix poorly by moving rapidly to, and slowly away from, sections of the sample space. In this case, for example, the Markov chain may become entrapped within a mode of the probability function, with rare chance  of escaping to adequately mimic the occupation frequencies in the overall sample space. If modes of the unstable model are not unique, then important outcomes may be missed without multiple chains or impractically enormous numbers of MCMC samples. This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in any particulars of the MCMC algorithm, and similar complications can also arise for Metropolis-Hastings algorithms for MCMC. Hence, while an unstable FOES model $P_{\boldsymbol \theta_N}$ may be valid and technically open to simulation by MCMC, the aspect of instability can render applications of MCMC as numerically infeasible for simulation purposes. This result is in line with conclusions of @handcock2003assessing and @schweinberger2011instability for other exponential models.

@@

The potential numerical difficulties with maximum likelihood with unstable models, as described in the previous section, can naturally carry over to Bayes inference. Considering that the degeneracy issues related to unstable models can cause likelihoods can be flat (e.g., near zero) for many parameters under a given data outcome and that sign changes in parameters can shift tremendous probability to extreme and opposite outcomes in the sample space (e.g., Corollary \@ref(cor:sign), Theorem \@ref(thm:something)), then numerical complications may arise with Bayes inference in sampling a posterior parameter space based on MCMC. The potential challenges in chain mixing are similar to those presented in Section \@ref(mcmc), though in chain movements through the parameter space as opposed to the sample space in data generation. That is, in the Bayes setting for sampling a posterior distribution for $\boldsymbol \theta_N$,  a chain may  unstable to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one-component changes in the sample space may impact data simulation with unstable models. For example, if $\pi(\cdot)$ denotes a prior density for $\boldsymbol \theta_N$ and $q(\cdot | \cdot)$  denotes a proposal distribution  for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes 
$$
\alpha\left(\boldsymbol \theta^{(1)} \mid \boldsymbol \theta^{(2)}\right)= \min\left\{1,
\frac{q(\boldsymbol \theta_N^{(2)} \mid \boldsymbol \theta_N^{(1)})}{q(\boldsymbol \theta_N^{(1)} \mid \boldsymbol \theta_N^{(2)})}
\frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
$$
which indicates how parameter sensitivity in the likelihood $P_{\boldsymbol \theta_N} (  \boldsymbol x)$ may complicate sampling of the posterior $P_{\boldsymbol \theta_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N)$ (i.e., moving from $\boldsymbol \theta^{(1)}$ to $\boldsymbol \theta^{(2)}$ in the parameter space). Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \@ref(rbm)-\@ref(deep-learning). As latent variables are often sampled with parameters in a Bayes MCMC approach, this aspect may further compound numerical problems in chain mixing.
